Data 621---HW1---Regression Model---Version 2

### 2020-09-09 HW1/RegModel_Adler.Rmd
Refactor to build a 75/25 train/test split and proceed

@@

```{r imputeMean}
MBmelt[, Value := replace(Value, is.na(Value), mean(Value, na.rm = TRUE)),
       keyby = Metric]
```

@@

```{r centerScale}
MBmeltSC <- copy(MBmelt)
MBmeltUnscale <- MBmeltSC[, .(Mean = mean(Value), SD = sd(Value)),
                          keyby = Metric]
MBmeltSC[, Value := scale(Value, center = TRUE, scale = TRUE), keyby = Metric]
knitr::kable(MBmeltUnscale, digits = 2L, caption = 'Data Means and SDs')
```
```{r plotScaled, fig.width=7}
ggplot(MBmelt, aes(x = Value)) + geom_density() +
  facet_wrap(~ Metric, scales = 'free') +
  ggtitle(label = 'Kernel-Smoothed Densities of Raw Data')
ggplot(MBmeltSC, aes(x = Value)) + geom_density() +
  facet_wrap(~ Metric, scales = 'free') +
  ggtitle(label = 'Kernel-Smoothed Densities of Centered and Scaled Data')
```

@@

BCL <- data.table(Metric = c('TEAM_PITCHING_H', 'TEAM_FIELDING_E',
                             'TEAM_BASERUN_SB', 'TEAM_PITCHING_HR',
                             'TEAM_PITCHING_BB', 'TEAM_PITCHING_SO',
                             'TEAM_BATTING_3B'),
                  Lambda = c(-2.5, -0.5, 0, 0.6, -0.15, -0.35, 0))
setkey(BCL, Metric)
knitr::kable(BCL, digits = 2L, caption = 'Hand-selected Box-Cox Lambdas')

# Left join MBmelt to BCL. 
MBmelt <- BCL[MBmelt]

# Transform data where there are selected lambdas
MBmelt[!is.na(Lambda), Value := BxCox(Value, Lambda)]

# Remove Lambda
MBmelt[, Lambda := NULL]

# Plot results
ggplot(MBmelt, aes(x = Value)) + geom_density() +
  facet_wrap(~ Metric, scales = 'free') +
  ggtitle(label = 'Kernel-Smoothed Densities of Box-Cox Transformed Data')
```

Most of the variables are now closer to a Gaussian distribution. Also note we
have removed `TEAM_BATTING_HR` from the data. Now the data needs to be rescaled
and recentered.

```{r reScale}
MBmeltSC <- copy(MBmelt)
MBmeltUnscale <- MBmeltSC[, .(Mean = mean(Value), SD = sd(Value)),
                          keyby = Metric]

MBmeltSC[, Value := scale(Value, center = TRUE, scale = TRUE), keyby = Metric]
knitr::kable(MBmeltUnscale, digits = 2L,
             caption = 'Post-Transform Data Means and SDs')
```

@@

```{r reImputeThenScale}
MBmeltSC <- copy(MBmelt)
MBmeltSC[, Lambda := NULL]
# Remove infinities
MBmeltSC[, Value := ifelse(is.infinite(Value), NA, Value)]
# Center and Scale
MBmeltSC[, Value := replace(Value, is.na(Value), mean(Value, na.rm = TRUE)),
       keyby = Metric]
MBmeltUnscale <- MBmeltSC[, .(Mean = mean(Value), SD = sd(Value)),
                          keyby = Metric]
MBmeltSC[, Value := scale(Value, center = TRUE, scale = TRUE), keyby = Metric]
knitr::kable(MBmeltUnscale, digits = 2L,
             caption = 'Final Post-Transform Data Means and SDs')
ggplot(MBmeltSC, aes(x = Value)) + geom_density() +
  facet_wrap(~ Metric, scales = 'free') +
  ggtitle(label = 'Kernel-Smoothed Densities of Scaled Box-Cox Transformed Data')
```

@@

## Analysis
Normally, we would use a statistical packahe like `caret`, which not only
simplifies model construction but also takes care of cross-validation. For the
purposes of this homework, and to demonstrate understanding of the underlying
decisions involved in predictive modeling, we will use calls to base R for the
null and saturated models and the `stepAIC` function from the `MASS` package to
perform a stepwise linear regression using AIC as our goodness of fit metric.
In the final section, we may pivot to using `caret` for final model selection.

@@

### Linear Models
We will show three models: the null model, the saturated model, and the model
resulting from the stepwise regression. For the regression, we will consider at
most pairwise interactions within the classes of metrics.

```{r lm, include=FALSE}
MBsc <- dcast(MBmeltSC, INDEX ~ Metric, value.var = 'Value')
# Replace TARGET_WINS with original. Could have restored it from MBmeltUnscale.
MBsc[, TARGET_WINS := MB[, TARGET_WINS]]
# Removing INDEX for convenience. Could have used "-INDEX" in the formula call.
MBsc[, INDEX := NULL]

nullModel <- lm(TARGET_WINS ~ 1, data = MBsc)
snm <- summary(nullModel)
nmr2 <- prettyNum(snm$r.squared, digits = 3L)
nmar2 <- prettyNum(snm$adj.r.squared, digits = 3L)
nmAIC <- prettyNum(AIC(nullModel), digits = 2L, big.mark = ',')

saturatedModel <- lm(TARGET_WINS ~ ., data = MBsc)
ssm <- summary(saturatedModel)
smr2 <- prettyNum(ssm$r.squared, digits = 3L)
smar2 <- prettyNum(ssm$adj.r.squared, digits = 3L)
smAIC <- prettyNum(AIC(saturatedModel), digits = 2L, big.mark = ',')

stepModel <- stepAIC(saturatedModel, scope = list(upper = ~
  (TEAM_BASERUN_CS + TEAM_BASERUN_SB) ^ 2 + (TEAM_BATTING_2B + TEAM_BATTING_3B +
  TEAM_BATTING_BB + TEAM_BATTING_H + TEAM_BATTING_HR + TEAM_BATTING_SO) ^ 2 +
  (TEAM_FIELDING_DP + TEAM_FIELDING_E) ^ 2 + (TEAM_PITCHING_BB +
  TEAM_PITCHING_H + TEAM_PITCHING_HR + TEAM_PITCHING_SO) ^ 2, lower = ~ 1),
  direction = 'both')

sstm <- summary(stepModel)
stmr2 <- prettyNum(sstm$r.squared, digits = 3L)
stmar2 <- prettyNum(sstm$adj.r.squared, digits = 3L)
stmAIC <- prettyNum(AIC(stepModel), digits = 2L, big.mark = ',')

```
```{r nullOutput}
knitr::kable(snm$coefficients, digits = 3L,
             caption = 'Null Model Output')
```
The null model has an \(R^2\) of `r nmr2`, an adjusted \(R^2\) of `r nmar2` and
an AIC of `r nmAIC`.

```{r satOutput}
knitr::kable(ssm$coefficients, digits = 3L,
             caption = 'Saturated Model Output')
```
The saturated model has an \(R^2\) of `r smr2`, an adjusted \(R^2\) of `r smar2`
and an AIC of `r smAIC`.

@@

The step model allowing pairwise interactions within the four classes of
metric: base running, batting, fielding, and pitching, has an \(R^2\) of
`r stmr2`, an adjusted \(R^2\) of `r stmar2` and an AIC of `r stmAIC`.

### Discussion
The first item for discussion is the observation that the intercept does **not**
change between the null (intercept-only) model and the saturated (all variables
included) model. There are 162 games in a baseball season. According to both the
null and saturated models, given no other information, a team will *lose*
slightly more than half its games. Whereas in the stepwise model allowing for
pairwise interactions within classes, given no other information, a team will
*win* slightly more than half of its games. Apparently, inteactions unearth
somewhat more information about why teams **lose** rather than why they **win**.

@@

 * Reasonable
   * Caught stealing removes baserunners; negative coefficient makes sense.
   * Stolen bases get a runner closer to home plate; positive coefficient makes
   sense.
   * Triples get a runner very close to home plate; positive coefficient makes
   sense.
   * Walks are free baserunners; positive coefficient makes sense.
   * Hits increase the number of baserunners; positive coefficient makes sense.
   * Hitting home runs directly increase the score; positive coefficient makes
   sense.
   * Striking out reduces the number of baserunners; negative coefficient makes
   sense.
   * Errors allow the other team free baserunners; negative coefficient makes
   sense.
   * Giving up home runs gives the opponent scores; negative coefficient makes
   sense.
   * Getting strikeouts reduces the opponents baserunners; positive coefficient
   makes sense.
 * Curious
   * Hitting doubles get runners on base; why negative?
   * Turning double plays reduce baserunners; why negative?
   * Giving up walks allows the other team free baserunners; its coefficient is
   both insignificant and near 0.
   * Giving up hits allows the other team baserunners; its coefficient is
   both insignificant and near 0.

An absolutely fascinating observation. In the first draft of this exercise,
`TEAM_BATTING_HR` was removed due to its high correlation with
`TEAM_PITCHING_HR`. Note the coefficients, batting is +10 and pitching is -7.
However, this is such as strong indicator, that in the first run, pitching was
given a *postive* coefficient of around +3. In hindsight this is because
*it was being used as an indicator for* ***batting!!*** The correlation allowed
the use of pitching HRs as an indicator for the hidden batting HRs! Once both
were restored to the model, the logical coefficients surfaced. Another reason
why models should not be trusted out of the box, but all model results should
be reviewed for sanity and sense!!

The curiosities above can *possibly* be resolved by looking at the interaction
terms. When giving up walks is interacted with strikeouts,
`TEAM_PITCHING_BB:TEAM_PITCHING_SO`, the coefficent is positive and significant.
Perhaps pitchers with many strikeouts also walk a lot of batters so giving up
walks on its own is less relevant. Similarly, the interaction between giving
up hits and strikeouts, `TEAM_PITCHING_H:TEAM_PITCHING_SO `, is highly negative
and significant. Once again, it probably captures most of the giving up hits
information making the singleton less relevant.

A possible explanation for the negative coefficent for getting double plays, is
that double plays require at least two people on base. That means that the
oppenent has a lot of base runners, which is very highly correlated with
scoring.

@@

>>>>

### Training and Test Sets
As we are performing this analysis manually, we will break up the supplied data
into two sets, one for training and one for testing. This needs to be done
before any normalisation or imputation, otherwise the testing set is polluted
with information from the training set. Please note that obervations prior to
this point were on the entirety of the data set.

```{r testTrain}
set.seed(811247)
trainObs <- sample(x = n, size = 0.75 * n, replace = FALSE)
MBtrn <- MB[trainObs]
MBtst <- MB[!trainObs]
MBtrnlng <- melt(MBtrn, id.vars = 'INDEX', variable.name = 'Metric',
                 value.name = 'Value', variable.factor = FALSE)
MBtstlng <- melt(MBtst, id.vars = 'INDEX', variable.name = 'Metric',
                 value.name = 'Value', variable.factor = FALSE)

# Casting to double since medians and quartiles may return non-integer values.
MBtrnlng[, Value := as.double(Value)]
MBtstlng[, Value := as.double(Value)]
```

@@

```{r imputeMean}
MBtrnlng[, Value := replace(Value, is.na(Value), mean(Value, na.rm = TRUE)),
       keyby = Metric]
```

@@

```{r centerScale}
MBtrnlngSC <- copy(MBtrnlng)
MBtrnlngUnscale <- MBtrnlngSC[, .(Mean = mean(Value), SD = sd(Value)),
                              keyby = Metric]
MBtrnlngSC[, Value := scale(Value, center = TRUE, scale = TRUE), keyby = Metric]
knitr::kable(MBtrnlngUnscale, digits = 2L, caption = 'Data Means and SDs')
```
```{r plotScaled, fig.width=7}
ggplot(MBtrnlng, aes(x = Value)) + geom_density() +
  facet_wrap(~ Metric, scales = 'free') +
  ggtitle(label = 'Kernel-Smoothed Densities of Raw Data')
ggplot(MBtrnlngSC, aes(x = Value)) + geom_density() +
  facet_wrap(~ Metric, scales = 'free') +
  ggtitle(label = 'Kernel-Smoothed Densities of Centered and Scaled Data')
```

@@

BCL <- data.table(Metric = c('TEAM_PITCHING_H', 'TEAM_FIELDING_E',
                             'TEAM_BASERUN_SB', 'TEAM_PITCHING_HR',
                             'TEAM_BATTING_HR', 'TEAM_PITCHING_BB',
                             'TEAM_PITCHING_SO', 'TEAM_BATTING_3B',
                             'TEAM_BASERUN_CS'),
                  Lambda = c(-2.5, -0.7, 0, 0.6, 0.6, -0.1, -0.35, 0, 0.5))
setkey(BCL, Metric)
knitr::kable(BCL, digits = 2L, caption = 'Hand-selected Box-Cox Lambdas')

# Left join MBtrnlng to BCL. 
MBtrnlng <- BCL[MBtrnlng]

# Transform data where there are selected lambdas
MBtrnlng[!is.na(Lambda), Value := BxCox(Value, Lambda)]

# Remove Lambda
MBtrnlng[, Lambda := NULL]

# Plot results
ggplot(MBtrnlng, aes(x = Value)) + geom_density() +
  facet_wrap(~ Metric, scales = 'free') +
  ggtitle(label = 'Kernel-Smoothed Densities of Box-Cox Transformed Data')
```

Most of the variables are now closer to a Gaussian distribution. Now the data
needs to be rescaled and recentered.

```{r reScale}
MBtrnlngSC <- copy(MBtrnlng)
MBtrnlngUnscale <- MBtrnlngSC[, .(Mean = mean(Value), SD = sd(Value)),
                          keyby = Metric]

MBtrnlngSC[, Value := scale(Value, center = TRUE, scale = TRUE), keyby = Metric]
knitr::kable(MBtrnlngUnscale, digits = 2L,
             caption = 'Post-Transform Data Means and SDs')
```

@@

```{r reImputeThenScale}
MBtrnlngSC <- copy(MBtrnlng)

# Remove infinities
MBtrnlngSC[, Value := ifelse(is.infinite(Value), NA, Value)]
# Center and Scale
MBtrnlngSC[, Value := replace(Value, is.na(Value), mean(Value, na.rm = TRUE)),
       keyby = Metric]
MBtrnlngUnscale <- MBtrnlngSC[, .(Mean = mean(Value), SD = sd(Value)),
                          keyby = Metric]
MBtrnlngSC[, Value := scale(Value, center = TRUE, scale = TRUE), keyby = Metric]
knitr::kable(MBtrnlngUnscale, digits = 2L,
             caption = 'Final Post-Transform Data Means and SDs')
ggplot(MBtrnlngSC, aes(x = Value)) + geom_density() +
  facet_wrap(~ Metric, scales = 'free') +
  ggtitle(label = 'Kernel-Smoothed Densities of Scaled Box-Cox Transformed Data')
```

@@

## Analysis
Normally, we would use a statistical package like `caret`, which not only
simplifies model construction but also takes of splitting into training and
test sets and takes care of cross-validation. For the purposes of this homework,
and to demonstrate understanding of the underlying decisions involved in
predictive modeling, we will use calls to base R and the `MASS` package.

As the transforms were imposed on the entirety of the training data, we ar

A null model is once where there are no explanatory variables other than the
grand mean. A saturated model is one where every variable appears.

@@

### Linear Models
We will show three models:

 1. The null model: This is the model with only the intercept, the grand mean.
 2. The saturated model: this is the model with every term included
 3. A model allowing pairwise interactions within the data types of baserunning,
 batting, pitching, and fielding, fit using a forward and backwards stepwise
 regression, and the model
resulting from the stepwise regression. A nuFor the regression, we will consider at
most pairwise interactions within the classes of metrics.

```{r lm, include=FALSE}
MBtrnSC <- dcast(MBtrnlngSC, INDEX ~ Metric, value.var = 'Value')

# Rscale TARGET_WINS to its standard value.
MBtrnSC[, TARGET_WINS := TARGET_WINS * 
          MBtrnlngUnscale[Metric == 'TARGET_WINS', SD] +
          MBtrnlngUnscale[Metric == 'TARGET_WINS', Mean]]

# Removing INDEX for convenience. Could have used "-INDEX" in the formula call.
MBtrnSC[, INDEX := NULL]

nullModel <- lm(TARGET_WINS ~ 1, data = MBtrnSC)
snm <- summary(nullModel)
nmr2 <- prettyNum(snm$r.squared, digits = 3L)
nmar2 <- prettyNum(snm$adj.r.squared, digits = 3L)
nmAIC <- prettyNum(AIC(nullModel), digits = 2L, big.mark = ',')
nmRMSE <- prettyNum(sqrt(mean(nullModel$residuals ^ 2)), digits = 5L)

saturatedModel <- lm(TARGET_WINS ~ ., data = MBtrnSC)
ssm <- summary(saturatedModel)
smr2 <- prettyNum(ssm$r.squared, digits = 3L)
smar2 <- prettyNum(ssm$adj.r.squared, digits = 3L)
smAIC <- prettyNum(AIC(saturatedModel), digits = 2L, big.mark = ',')
smRMSE <- prettyNum(sqrt(mean(saturatedModel$residuals ^ 2)), digits = 5L)
smF <- prettyNum(ssm$fstatistic, digits = 5L)
smFp <- pf(ssm$fstatistic[[1]], ssm$fstatistic[[2]], ssm$fstatistic[[3]],
            lower.tail = FALSE)

crazyModel <- lm(TARGET_WINS ~
  (TEAM_BASERUN_CS + TEAM_BASERUN_SB) ^ 2 + (TEAM_BATTING_2B + TEAM_BATTING_3B +
  TEAM_BATTING_BB + TEAM_BATTING_H + TEAM_BATTING_HR + TEAM_BATTING_SO) ^ 2 +
  (TEAM_FIELDING_DP + TEAM_FIELDING_E) ^ 2 + (TEAM_PITCHING_BB +
  TEAM_PITCHING_H + TEAM_PITCHING_HR + TEAM_PITCHING_SO) ^ 2, data = MBtrnSC)

stepModel <- stepAIC(crazyModel, scope = list(upper = ~
  (TEAM_BASERUN_CS + TEAM_BASERUN_SB) ^ 2 + (TEAM_BATTING_2B + TEAM_BATTING_3B +
  TEAM_BATTING_BB + TEAM_BATTING_H + TEAM_BATTING_HR + TEAM_BATTING_SO) ^ 2 +
  (TEAM_FIELDING_DP + TEAM_FIELDING_E) ^ 2 + (TEAM_PITCHING_BB +
  TEAM_PITCHING_H + TEAM_PITCHING_HR + TEAM_PITCHING_SO) ^ 2, lower = ~ 1),
  direction = 'both')

sstm <- summary(stepModel)
stmr2 <- prettyNum(sstm$r.squared, digits = 3L)
stmar2 <- prettyNum(sstm$adj.r.squared, digits = 3L)
stmAIC <- prettyNum(AIC(stepModel), digits = 2L, big.mark = ',')
stmRMSE <- prettyNum(sqrt(mean(stepModel$residuals ^ 2)), digits = 5L)
stmF <- prettyNum(sstm$fstatistic, digits = 5L)
stmFp <- pf(sstm$fstatistic[[1]], sstm$fstatistic[[2]], sstm$fstatistic[[3]],
            lower.tail = FALSE)
```
```{r nullOutput}
knitr::kable(snm$coefficients, digits = 3L,
             caption = 'Null Model Output')
```
The null model has an \(R^2\) of `r nmr2`, an adjusted \(R^2\) of `r nmar2`, an
AIC of `r nmAIC`, and an RMSE on the training set of `r nmRMSE`. There is no
F-statistic for the entirety of the null model as there is only one axis for
degrees of freedom.

```{r satOutput}
knitr::kable(ssm$coefficients, digits = 3L,
             caption = 'Saturated Model Output')
```
The saturated model has an \(R^2\) of `r smr2`, an adjusted \(R^2\) of
`r smar2`, an AIC of `r smAIC`, and an RMSE on the training set of `r smRMSE`.
The p-value for the F-test with a critical value of `r smF[[1]]` with
`r smF[[2]]` and `r smF[[3]]` degrees of freedom is below machine error
(`r smFp`).

@@

The step model allowing pairwise interactions within the four classes of
metric: base running, batting, fielding, and pitching, has an \(R^2\) of
`r stmr2`, an adjusted \(R^2\) of `r stmar2`, an AIC of `r stmAIC`, and an RMSE
on the training set of `r stmRMSE`. The p-value for the F-test with a critical
value of `r stmF[[1]]` with `r stmF[[2]]` and `r stmF[[3]]` degrees of freedom
is below machine error (`r stmFp`).

### Discussion
The first item for discussion is the observation that while the intercept does
not change between the null model and the saturated model, it does decrease for
the model allowing within class pairwise interactions. There are 162 games in a
baseball season. According to all the models, given no other information, a team
will lose slightly more than half its games. Apparently, interctions in the
stepwise model unearth somewhat more information about why teams win.

@@

 * Reasonable
   * Caught stealing removes baserunners; negative coefficient makes sense.
   * Stolen bases get a runner closer to home plate; positive coefficient makes
   sense.
   * Triples get a runner very close to home plate; positive coefficient makes
   sense.
   * Walks are free baserunners; positive coefficient makes sense.
   * Hits increase the number of baserunners; positive coefficient makes sense.
   * Hitting home runs directly increase the score; positive coefficient makes
   sense.
   * Striking out reduces the number of baserunners; negative coefficient makes
   sense.
   * Errors allow the other team free baserunners; negative coefficient makes
   sense.
   * Giving up home runs gives the opponent scores; negative coefficient makes
   sense.
   * Getting strikeouts reduces the opponents baserunners; positive coefficient
   makes sense.
 * Curious
   * Hitting doubles get runners on base; why negative?
   * Turning double plays reduce baserunners; why negative?
   * Giving up walks allows the other team free baserunners; negative
   coefficient makes sense but why insignificant?
   * Giving up hits allows the other team baserunners; why positive athough
   insignificant?

An absolutely fascinating observation. In the first draft of this exercise,
`TEAM_BATTING_HR` was removed due to its high correlation with
`TEAM_PITCHING_HR`. Note the coefficients, batting is almost +12 and pitching is
-9. However, this is such as strong indicator, that in the first run, pitching
was given a *postive* coefficient of around +3. In hindsight this is because
*it was being used as an indicator for* ***batting!!*** The correlation allowed
the use of pitching HRs as an indicator for the hidden batting HRs! Once both
were restored to the model, the logical coefficients surfaced. Another reason
why models should not be trusted out of the box, but all model results should
be reviewed for sanity and sense!!

The curiosities above can *possibly* be resolved by looking at the interaction
terms. The interaction between giving up hits and strikeouts,
`TEAM_PITCHING_H:TEAM_PITCHING_SO `, is highly negative and significant. It
probably captures most of the giving up hits information making the singleton
less relevant.

A possible explanation for the negative coefficent for getting double plays, is
that double plays require at least two people on base. That means that the
opponent has a lot of base runners, which is very highly correlated with
scoring.

@@

General inferences about the nature of the coefficients is in the previous
section. The fascinating example of how multi-collinearity can override logic is
also discussed in the previous secton.

In order to make inferences from the model, the incoming data needs to be
processed in the exact same way:

 1. Mean imputation
 2. Box-Cox with the selected lambdas
 3. Mean imputation again to remove `Inf` issues
 4. Centering and scaling

### Model Evaluations
In the table below are the RMSE, \(R^2\), and significance of the F-statistic
for each of the three models as applied to the **testing** set.

```{r buildTests}
MBtstlng <- BCL[MBtstlng]
MBtstlng[!is.na(Lambda), Value := BxCox(Value, Lambda)]
MBtstlng[, Lambda := NULL]
MBtstlngSC <- copy(MBtstlng)

# Remove infinities
MBtstlngSC[, Value := ifelse(is.infinite(Value), NA, Value)]

# Center and Scale
MBtstlngSC[, Value := replace(Value, is.na(Value), mean(Value, na.rm = TRUE)),
       keyby = Metric]
MBtstlngUnscale <- MBtstlngSC[, .(Mean = mean(Value), SD = sd(Value)),
                          keyby = Metric]
MBtstlngSC[, Value := scale(Value, center = TRUE, scale = TRUE), keyby = Metric]
MBtstSC <- dcast(MBtstlngSC, INDEX ~ Metric, value.var = 'Value')
MBtstSC[, TARGET_WINS := TARGET_WINS * 
          MBtstlngUnscale[Metric == 'TARGET_WINS', SD] +
          MBtstlngUnscale[Metric == 'TARGET_WINS', Mean]]
```
